---
title: "Get started with CellSight"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with CellSight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

CellSight turns a single-cell or spatial object into a self-contained Shiny
app. You do not write any Shiny code: you describe *what to show* in a config
object, and CellSight writes both the app's data files and the app's source.

The code in this article is not evaluated when the site is built, because
building an app needs a real dataset. Run it against your own object.

## The two-step workflow

Every CellSight app is built in two steps, in this order:

1. **`makeShinyFiles()`** reads your object and writes the app's *data* into
   `shiny.dir` — chunked HDF5 expression matrices plus `.rds` files holding the
   config and cell metadata, all prefixed with `shiny.prefix` (`sc1` by
   default).
2. **`makeShinyCodes()`** writes the app's *code* into the same directory —
   `server.R`, `ui.R` and `shinyFunc.R`.

The split matters because the data step is the slow one, and because the prefix
is what ties the two together. Whatever you pass as `shiny.prefix` to
`makeShinyFiles()` must be passed again to `makeShinyCodes()`.

## A minimal app

Using the [multimodal PBMC CITE-seq
object](https://zenodo.org/records/15162323/files/multimodal_pbmc.rds?download=1)
(162,000 cells, 228 antibodies):

```{r}
library(Seurat)
library(cellsight)

seu <- readRDS("multimodal_pbmc.rds")

scConf <- createConfig(seu)

makeShinyFiles(seu, scConf, shiny.prefix = "sc1", shiny.dir = "shinyApp/")

makeShinyCodes(
  shiny.title  = "PBMC multiomics",
  shiny.prefix = "sc1",
  shiny.dir    = "shinyApp/"
)
```

That leaves a complete app in `shinyApp/`. Open `shinyApp/ui.R` in RStudio and
click **Run App**, or from the console:

```{r}
shiny::runApp("shinyApp")
```

See `vignette("deployment")` for hosting it somewhere other than your laptop.

## The config object

`createConfig()` inspects your object's cell metadata and returns a
`data.table` — conventionally called `scConf` — with one row per metadata
column it will expose in the app. It decides which columns are usable,
picks a colour palette for each categorical one, and guesses which two should
be shown by default.

```{r}
scConf <- createConfig(seu)
showOrder(scConf)
```

You will usually want to adjust it: rename metadata for display, drop
uninteresting columns, fix the colours. That is what the `mod*` helpers are
for, and it is worth doing before you build — see
`vignette("configuration")`.

Two arguments are worth knowing up front:

```{r}
# Only expose specific metadata columns
scConf <- createConfig(seu, meta.to.include = c("orig.ident", "seurat_clusters", "celltype"))

# Allow more categories per metadata column (default 50; anything above is dropped)
scConf <- createConfig(seu, maxLevels = 100)
```

`createConfig()` errors if it finds no multi-level metadata at all, since an
app with nothing to group cells by would be empty. If you hit that, your object
probably has not been clustered yet.

Before building, check the config still matches the object:

```{r}
checkConfig(scConf, seu)
```

## Setting sensible defaults

The app opens on whichever metadata and genes you nominate. Setting these saves
every future visitor a few clicks:

```{r}
scConf <- modDefault(scConf, "celltype", "seurat_clusters")

makeShinyFiles(
  seu, scConf,
  shiny.prefix      = "sc1",
  shiny.dir         = "shinyApp/",
  default.gene1     = "CD3E",
  default.gene2     = "MS4A1",
  default.multigene = c("CD3E", "MS4A1", "CD14", "NKG7", "PPBP"),
  default.dimred    = "umap"
)
```

`default.dimred` is also how you choose between reductions when your object has
several; `dimred.to.use` restricts which ones are written at all.

## Beyond gene expression

`makeShinyFiles()` looks at the object you hand it and writes every modality it
finds — spatial images, ATAC coverage, precomputed DEG tables — without you
calling anything extra. The details of what triggers each one, and the arguments
each needs, are in `vignette("modalities")`.

## Several datasets in one app

Pass a vector of prefixes to `makeShinyCodes()`, one per `makeShinyFiles()`
call, plus a header for each. Each dataset's tabs are grouped under its own
navbar dropdown:

```{r}
makeShinyFiles(seu1, scConf1, shiny.prefix = "sc1", shiny.dir = "shinyApp/")
makeShinyFiles(seu2, scConf2, shiny.prefix = "sc2", shiny.dir = "shinyApp/")

makeShinyCodes(
  shiny.title   = "Two-dataset atlas",
  shiny.prefix  = c("sc1", "sc2"),
  shiny.headers = c("Healthy", "Disease"),
  defPtSiz      = c(1.25, 1.5),
  shiny.dir     = "shinyApp/"
)
```

`shiny.headers` is ignored for a single dataset, but its length must match
`shiny.prefix` when you have more than one. `defPtSiz` takes either one value
for all datasets or one per dataset — reach for a smaller value when a dataset
has enough cells that the default overplots.

## Footnotes and citations

`shiny.footnotes` accepts either a string or a named list, which is rendered as
a formatted citation in the app's footer:

```{r}
makeShinyCodes(
  shiny.title = "PBMC multiomics",
  shiny.footnotes = list(
    author  = "Liu X., Ouyang J.F., Rossello F.J. et al.",
    title   = "",
    journal = "Nature",
    volume  = "586",
    page    = "101-107",
    year    = "2020",
    doi     = "10.1038/s41586-020-2734-6",
    link    = "https://www.nature.com/articles/s41586-020-2734-6"
  ),
  shiny.prefix = "sc1",
  shiny.dir    = "shinyApp/"
)
```

## Where to go next

* `vignette("configuration")` — shape the config: labels, colours, ordering.
* `vignette("modalities")` — spatial, scATAC-seq, DEG and `.h5ad` inputs.
* `vignette("deployment")` — run the app somewhere other than your laptop.
* `vignette("docker")` — the prebuilt container and its helper scripts.
* `vignette("templates")` — change what the generated app itself contains.
