---
title: "Editing the frontend templates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Editing the frontend templates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This article is for contributors changing what a generated CellSight app
*contains*. If you only want to change what your app shows, the config helpers
in `vignette("configuration")` are the right tool and you do not need any of
this.

CellSight's frontend is generated, not hand-written.
[`makeShinyCodes()`](../reference/makeShinyCodes.html) assembles `server.R`,
`ui.R` and `shinyFunc.R` from [jinjar](https://davidchall.github.io/jinjar/)
templates shipped in `inst/templates`. The rule that follows from that: **edit
the template, never the generated file.** Anything you change in a generated
`ui.R` is lost the next time anyone rebuilds.

## The workflow

1. Edit the template that owns the piece you are changing (see the map below).
2. Reinstall CellSight so the updated templates are on disk — they are read from
   the *installed* package, not your source tree.
3. Regenerate an app with your normal `makeShinyFiles()` + `makeShinyCodes()`
   call.
4. Run `Rscript tools/render-and-lint.R` before you commit.

Step 2 is the one people skip. `makeShinyCodes()` reads
`system.file("templates", package = "cellsight")`, so editing
`inst/templates/ui.R.jinja` in your checkout has no effect until you reinstall:

```{r}
devtools::install(".")
# or: install.packages(".", repos = NULL, type = "source")
```

## Layout

```
templates/
  server.R.jinja          master: orders the server.R blocks
  ui.R.jinja              master: orders the ui.R blocks (+ single/multi layout)
  shinyFunc.R.jinja       self-contained: the full shinyFunc.R
  partials/
    server/…              one file per server.R block
    ui/…                  one file per ui.R block
```

The division of labour is what tells you which file to open:

* The **masters** (`server.R.jinja`, `ui.R.jinja`) hold orchestration — which
  blocks appear, in what order, gated on whether a dataset has spatial / ATAC /
  DEG data, and the single- versus multi-dataset layout. Edit a master to change
  *which* blocks appear or their order.
* A **partial** holds the content of one block. Edit a partial to change what
  one tab does.
* `shinyFunc.R.jinja` is essentially a plain R file of shared plotting
  functions, with one trailing
  `{% for d in datasets %}{% if d.has_bw %}…{% endif %}{% endfor %}` block that
  appends the ATAC track-plot helpers once per ATAC dataset. Edit it as ordinary
  R and leave that trailing block in place.

## Partial reference

Each interactive tab is one partial in both `server/` (reactive logic) and `ui/`
(layout), named for the tab's purpose:

| file (server/ + ui/)  | tab title in the app |
| --------------------- | -------------------- |
| `dimred-zoom`         | Zoom-enable Dimred   |
| `dimred-sidebyside`   | Side-by-side DimRed  |
| `dimred-coexpression` | Gene coexpression    |
| `dimred-split`        | Split-out DimRed     |
| `deg-dimred`          | DEG dimred plot      |
| `spatial-zoom`        | Zoom-enable Spatial  |
| `spatial-sidebyside`  | Side-by-side Spatial |
| `trackplot`           | Track plot (ATAC)    |
| `violin-box`          | Violinplot / Boxplot |
| `proportion`          | Proportion plot      |
| `bubble-heatmap`      | Bubbleplot / Heatmap |
| `module-score`        | Module scoring       |

### Spatial layout variants

The spatial tabs need variants that the masters choose between:

| file | when it is used |
| ---- | --------------- |
| `spatial-zoom` / `spatial-sidebyside` | single-slide dataset |
| `spatial-zoom-multislide` / `spatial-sidebyside-multislide` | multi-slide dataset; adds a slide selector |
| `spatial-sidebyside-extra` | **unused.** A legacy duplicate of `spatial-sidebyside`, no longer included after the menus were regrouped. Kept for reference only — do not add to it |

### Blocks with no partial

Short, structural or rarely-edited blocks live directly in the masters:

| block | where it lives |
| ----- | -------------- |
| library calls | top of `server.R.jinja` and `ui.R.jinja` |
| preamble (palettes, panel sizes, `shinyServer(...)` / `shinyUI(...)` open, helpers) | after the library calls in each master |
| per-dataset data loads | the first `{% for d in datasets %}` loop in each master, with `-spatial` / `-track` variants gated on `d.has_image` / `d.has_bw` |
| closing lines | end of `server.R.jinja` and `ui.R.jinja` |
| `google-analytics.html` | the `ga_html` string in `R/makeShinyCodes.R` |

## Template syntax

The templates use **non-default jinjar delimiters**, because the generated R
code is full of its own `{` and `}` that must pass through untouched:

* `<< var >>` — variable interpolation (jinjar's usual `{{ }}` is remapped)
* `{% if %}` / `{% for %}` / `{% include %}` — control tags (jinjar defaults)

That remapping is done by `scJinjaCfg()`, so any new rendering code should use
it rather than a bare `jinjar::jinjar_config()`.

## The template context

Everything the templates can see is built in R, in `makeShinyCodes()`. Inside a
`{% for d in datasets %}` loop the per-dataset fields are `<< d.prefix >>`,
`<< d.ptsiz >>` and so on. The fields available on each `d`:

| field | meaning |
| ----- | ------- |
| `prefix` | the dataset's `shiny.prefix` |
| `header` | its navbar header (multi-dataset layouts) |
| `ptsiz` | default point size |
| `has_image` | spatial data present |
| `has_bw` | ATAC bigWig data present |
| `has_deg` | precomputed DEG data present |
| `multislide` | more than one spatial slide |
| `slider_min` / `slider_max` / `slider_step` / `ptsiz_val` | spatial point-size slider parameters |
| `slider_*_first` / `ptsiz_val_first` | first-dataset variants, for the extra spatial tab in the multi-dataset layout |

The three `has_*` flags are **detected from disk**, not passed in:
`makeShinyCodes()` tests for `<prefix>image.rds`, `<prefix>bw.rds` and
`<prefix>deg.h5` in `shiny.dir`. So a tab that fails to appear usually means
`makeShinyFiles()` did not write that file — check the app directory before
suspecting the template.

If you need a new fact in a template, add it to the `datasets` list in
`R/makeShinyCodes.R` first; templates cannot compute it themselves.

## Formatting

Do **not** try to keep a partial's indentation "correct" for the final app. A
partial is `{% include %}`d at different nesting depths — a tab sits inside a
content `navbarMenu(...)` in both layouts, and one level deeper again in the
multi-dataset layout — so no single indentation could be right everywhere.

Instead, `makeShinyCodes()` runs [styler](https://styler.r-lib.org) on the
fully-assembled files, which re-indents and re-spaces the complete programs.
styler is formatting-only and optional: without it installed the generated files
are still valid, just un-styled.

Write partials to be readable on their own; leave final layout to styler.

### Whitespace still matters in the masters

styler tidies spacing but not structure, and jinjar emits any text *between*
control tags, newlines included. Keep the `{% … %}` tags in the masters on a
single line with no spaces between them — a stray newline between two
`{% include %}` tags injects a blank line into the output, and a structural line
break in the wrong place can change the assembled program.

## Linting and validation

Templates are linted **as rendered**, never as fragments: the fragments contain
jinjar tokens and are not valid R on their own. `tools/render-and-lint.R`
renders a fully-featured app — multi-dataset, spatial multi-slide, ATAC and DEG,
so every partial is exercised — then checks the output parses and lints it:

```bash
Rscript tools/render-and-lint.R
```

It is strict, and exits non-zero if a template fails to render, if the generated
R does not parse, or if the linter reports *any* finding. It runs on every pull
request via `.github/workflows/lint-templates.yml`, so a failure there is worth
reproducing locally first.

The lint configuration lets styler own layout (indentation off) and allows the
inherited house style — `=` assignment, camelCase and dotted names, explicit
`return()`. If you add a lint exclusion, do it in that script so CI and local
runs agree.

## Editor support

For syntax highlighting in mixed R + Jinja files, a VS Code extension is bundled
at `inst/extdata/vscode-rjinja/rjinja-syntax.tar.gz`. Extract it and install via
**Install from VSIX…**. It is optional and aimed at contributors editing
`*.R.jinja` files.
